Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2021-01-28 analysis/paper/paper.Rmd
mention separate models for CF and 1-ha
EB Edits to Intro

1.  Organisms experience multiple sources of natural and anthropogenic stress simultaneously. However, the impacts of multiple sources of stress on population dynamics are not well understood. Combined environmental stresses have the potential to interact to exacerbate or dampen additive effects on population dynamics. <!--# examples of interactions on demography -->For example, fire and herbivory ...[@tyeDemographicMenageTrois2016].

@@

2.  Habitat fragmentation is one such source of environmental stress. Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015]. In the Brazilian Amazon, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest farther than 1 km from an edge has increased to 75% compared to 90% historically [@haddadHabitatFragmentationIts2015]. Fragmentation can impact plant population dynamics through ... <!--# more about consequences of fragmentation pull from slides from Emilio-->

3.  Simultaneously, fragmented tropical forests are experiencing increased drought stress as a result of climate change. The northern Amazon has been experiencing a drying trend since the mid 1970s and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008]. Additionally, El Niño Southern Oscillation (ENSO)-induced droughts are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].

@@

4.  It has been hypothesized that drought and fragmentation could act synergistically to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999]. Fragmentation results in changes in forest microclimate near edges including increased solar radiation, increased temperature, and decreased humidity and soil moisture due to reduced microclimate buffering [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]. This decrease in microclimate buffering could exacerbate the effects of drought in fragments compared to intact forest. For example, drought-induced tree mortality has been observed to increase more near forest edges compared to interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].

5.  Alternatively, plant individuals that are more susceptible to desiccation may not have survived fragmentation, resulting in populations in fragments which are more resilient to drought.

@@

Second, there is a growing literature on how tropical plants respond to severe droughts, but few studies have compared the responses of plants in continuous forest with those in fragments. Laurance et al. [-@lauranceEffectsStrongDrought2001] found that the increase in tree (\>10 cm DBH) mortality during a drought were exacerbated in fragments, but the decline in mortality after the drought was also stronger in fragments. However, in a temperate system... <!--# oak paper --> [@forner2020].

@@

We investigated the effects of drought on the growth, survival, reproduction, and recruitment of an understory plant in an experimentally fragmented landscape. Specifically, we want to know 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest? 2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest? 3) Are the effects of drought and fragmentation on vital rates consistent across life-history stages?

@@

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW). In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig. -- <!--# fig ref -->). These fragments were isolated from 1980--1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried. The vegetation regenerating around fragments is cleared periodically to ensure their isolation [@bierregaardLessonsAmazoniaEcology2001]. The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m [@rankin-de-meronaPreliminaryResultsLargescale1992], and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999]. The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content. Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).

@@

>>>>

1.  Organisms simultanously experience multiple forms of natural and anthropogenic disturbance. While work investigating the demographic effects of individual factors is common, studies investigating how multiple factors interact to influence population dynamics remains poorly understood. They could act in concert to exacerbate or dampen population dynamics. <!--# examples of interactions on demography -->For example, fire and herbivory ...[@tyeDemographicMenageTrois2016].

@@

2.  Habitat fragmentation is a major source of environmental stress. **[I'm not sure I'd call it a stress...I think the biotic and abioti changes associated with it are stressful]**  Forest fragmentation, in particular, is widespread and has resulted in 70% of the Earth's remaining forest being within 1km of forest edge [@haddadHabitatFragmentationIts2015]. Of particualar concern are the effects of fragmentation on species-rich tropical ecosystems. In the Brazilian Amazon, for example, over 70,000 km of new forest edges are being created annually [@broadbentForestFragmentationEdge2008] and the proportion of forest >1 km from an forest edge has decreased to from 90% to only 75% [@haddadHabitatFragmentationIts2015]. Fragmentation can impact plant population dynamics through ... <!--# more about consequences of fragmentation pull from slides from Emilio-->

3.  In addition to the environmental changes associated with fragmentation, populations persisting in tropical forest remnants experience abiotic stress resulting from climate change. For instance, the northern Amazon has been experiencing a drying trend since the mid-1970s, and ensemble climate models predict substantial decreases in dry season precipitation in southern Amazonia for the 21st century [@malhi2008]. Furthermore, droughts induced by the El Niño Southern Oscillation (ENSO) are predicted to increase in frequency and severity [@caiIncreasingFrequencyExtreme2014].

@@

4.  It has been hypothesized that drought and fragmentation could act in synergy to further reduce population viability in fragmented landscapes, particularly in the tropics [@lauranceEffectsStrongDrought2001; @didhamEdgeStructureDetermines1999]. Fragmentation drives several changes in microclimate near forest edges, including increased air temperature, decreased relative humidity, reduced soil moisture, and increased penetration of solar radiation in the forest understory [@ewersFragmentationImpairsMicroclimate2013; @didhamEdgeStructureDetermines1999]. This decrease in microclimatic buffering could exacerbate the effects of drought in fragments compared to intact forest. This may be why the drought-induced mortality of trees is higher near forest edges than in forest interiors [@schwartz2019; @lauranceEffectsStrongDrought2001].

5.  Alternatively, the populations persisting in fragments may be more resilient to drought if individuals susceptible to desiccation died quickly after fragment isolation.

@@

Second, there is a growing literature on how tropical plants respond to droughts, but few studies have compared the responses of plants in continuous forest with those in forest fragments. Laurance et al. [-@lauranceEffectsStrongDrought2001] found that the increase in tree (\>10 cm DBH) mortality during a drought were exacerbated in fragments, but the decline in mortality after the drought was also stronger in fragments. However, in a temperate system... <!--# oak paper --> [@forner2020].

@@

We investigated the effects of drought on the growth, survival, reproduction, and recruitment of a understory herb in an experimentally fragmented landscape in the Central Amazon. Specifically, we asked 1) Does drought increase or decrease the growth, survival, and fertility rates of plant populations in continuous forest? 2) Are the effects of drought on the vital rates of populations in fragments similar in direction and magnitude to those in continuous forest? 3) Are the effects of drought and fragmentation on vital rates consistent across life-history stages?

@@

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW). In addition to large areas of continuous forest, the BDFFP has forest fragment reserves that range in size from 1-100 ha (Fig. -- <!--# fig ref -->). These fragments were isolated from 1980--1984 by felling the trees surrounding the area chosen for isolation and - in most cases - burning the downed trees once they dried. In subsequent decades the vegetation regenerating around fragments has been periodically cleared to ensure fragment isolation [@bierregaardLessonsAmazoniaEcology2001].

The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30--37m tall canopy, emergent trees up to 55m [@rankin-de-meronaPreliminaryResultsLargescale1992], and an understory dominated by stemless palms [@scariotForestFragmentationEffects1999]. The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content. Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).

@@

Because DLNMs do not currently allow modeling of interactions, we fit separate models for plants in fragments and in continuous forest to allow the shape of the crossbasis function to differ between habitats.
